---
title: "Marmap Demo"
author: "Adam A. Kemberling"
date: "2/8/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Marmap Package Deom

Lets see what this marine mapping package can do:

```{r}
library(marmap)
```

# From [Molecular Ecologist](http://www.molecularecologist.com/2015/07/marmap/)

Basic workflow:

 * Get bathymetry data from NOAA servers     
 * Create color pallettes that don't suck
 * Plot

```{r}
# Load package
library(marmap)


#  Fetch data on NOAA servers and write on disk
bat <- getNOAA.bathy(-20, 0, 40, 60, res = 4, keep=TRUE)



# Create nice looking color palettes
blues <- c("lightsteelblue4", "lightsteelblue3", "lightsteelblue2", "lightsteelblue1")
greys <- c(grey(0.6), grey(0.93), grey(0.99))



# Plot
plot(bat, image = TRUE, land = TRUE, lwd = 0.1, bpal = list(c(0, max(bat), greys), c(min(bat), 0, blues)))
plot(bat, lwd = 0.8, deep = 0, shallow = 0, step = 0, add = TRUE) # highlight coastline
```

## Spatial Analysis Features

Here, they used the existing R package gdistance (van Etten 2011) to develop an easy-to-use function with sensible defaults for marine ecologists. As an illustration, Viricel (2012) used least cost paths to infer the connectivity of spotted dolphins (Stenella frontalis) in the Gulf of Mexico. This species almost exclusively uses habitats located above the continental shelf and are virtually never found in waters deeper than 300m. Thus, these animals cannot cross the Gulf of Mexico from West to East along a straight line. To estimate potential gene flow (or lack thereof) between populations, computing accurate geographic distances between individuals must take into account the fact that individuals are restricted to the continental shelf when moving/migrating. The script used in the paper was similar to the following:


```{r}
# Import bathymetry
bat <- getNOAA.bathy(-100, -80, 22, 31, res = 1, keep = TRUE)

# Load location of individuals (these are NOT from Viricel 2012)
loc <- data.frame( x = c(-96.92707, -96.60861, -96.86875, -96.14351, -92.82518, -90.86053, -90.14208, -84.64081, -83.81274, -81.13277, -80.33498, -88.52732, -94.46049), y = c(25.38657, 25.90644, 26.57339, 27.63348, 29.03572, 28.16380, 28.21235, 26.71302, 25.12554, 24.50031, 24.89052, 30.16034, 29.34550) )

# Compute least cost paths between -5m and -300m. 
# Beware! Computation takes time with high resolution bathymetries!
tr <- trans.mat(bat, min.depth = -5, max.depth = -300)
cost <- lc.dist(tr, loc, res="path")

# Plot map with isobaths every 1000m
plot(bat, image = TRUE, land = TRUE, deep=-4000, shallow=-1000, step=1000, drawlabels = FALSE, bpal = list(c(min(bat,na.rm=TRUE), 0, blues), c(0, max(bat, na.rm=TRUE), greys)), lwd = 0.1)

# Add -300m isobath
plot(bat, deep = -300, shallow = -300, step = 0, lwd = 0.5, add = TRUE, drawlabels = TRUE)

# Add coastline
plot(bat, deep = 0, shallow = 0, step = 0, lwd = 1, add = TRUE)

# Add least cost paths and the position of individuals
dummy <- lapply(cost, lines, col = col2alpha("orange", 0.5), lwd = 0.8, lty = 1) 
points(loc, bg = "orange", cex = 0.8, pch = 21)
```

### Ggplot2 Support

```{r}
library(ggplot2)

# Load North West Atlantic dataset
data(nw.atlantic)
atl <- as.bathy(nw.atlantic)

# Plot with ggplot2
autoplot(atl, geom=c("raster", "contour"), colour="white", size=0.1) + 
  scale_fill_gradient2(low="dodgerblue4", mid="gainsboro", high="darkgreen")
```


### Buffer Areas

```{r}
# load and plot a bathymetry
data(florida)
plot(florida, lwd = 0.2)
plot(florida, n = 1, lwd = 0.7, add = TRUE)

# Create a point and a buffer around this point
loc <- data.frame(-80, 26)
buf <- create.buffer(florida, loc, radius=1.8)

# Get the surface within the buffer for several depth slices
surf1 <- get.area(buf, level.inf=-200, level.sup=-1)
surf2 <- get.area(buf, level.inf=-800, level.sup=-200)
surf3 <- get.area(buf, level.inf=-3000, level.sup=-800)

s1 <- round(surf1$Square.Km)
s2 <- round(surf2$Square.Km)
s3 <- round(surf3$Square.Km)

# Add buffer elements on the plot
col.surf1 <- rgb(0.7, 0.7, 0.3, 0.3)
col.surf2 <- rgb(0, 0.7, 0.3, 0.3)
col.surf3 <- rgb(0.7, 0, 0, 0.3)

plotArea(surf1, col = col.surf1)
plotArea(surf2, col = col.surf2)
plotArea(surf3, col = col.surf3)
plot(outline.buffer(buf), add = TRUE, lwd = 0.7)
points(loc, pch = 19, col = "red")

## Add legend
legend("topleft", fill = c(col.surf1, col.surf2, col.surf3),
       legend = c(paste("]-200 ; -1] -",s1,"km2"),
                  paste("]-800 ; -200] -",s2,"km2"),
                  paste("]-3000 ; -800] -",s3,"km2")))
```




### Tracking movement and depth using gps coordinates

```{r, eval = FALSE}
head(track1)

head(track2)

# Download bathymetric data and save on disk
bat <- getNOAA.bathy(174, 176, -37, -35, res = 1, keep = TRUE)

# Get depth profile along both tracks and remove positive depths (since random fake values can be on land)
# path.profile() gets depth value for all cells of the bathymetric grid below the gps tracks
path1 <- path.profile(track1[,-3], bat) ; path1 <- path1[-path1[,4]>0,]
path2 <- path.profile(track2[,-3], bat) ; path2 <- path2[-path2[,4]>0,]

# Get depth values for each gps tracking point
# get.depth() retrieve depth values only for gps tracking points
depth1 <- get.depth(bat, track1$lon, track1$lat, locator = FALSE, distance = TRUE) ; 
depth2 <- get.depth(bat, track2$lon, track2$lat, locator = FALSE, distance = TRUE) ; 

# Add depth values to tracks 1 and 2 and remove positive depths
track1$depth <- depth1$depth ; track1 <- track1[-track1$depth > 0,]
track2$depth <- depth2$depth ; track2 <- track2[-track2$depth > 0,]

# Plot
layout(matrix(c(1, 2, 1, 3, 4, 4), ncol = 2, byrow = TRUE), height = c(1, 1, 1))

## Bathymetric map with gps tracks
plot(bat, land = TRUE, image = TRUE, lwd = 0.2, bpal = list(c(min(bat,na.rm=TRUE), 0, blues), c(0, max(bat, na.rm=TRUE), greys)), ylim=c(-36.5,-35.25))
plot(bat, deep = 0, shallow = 0, step = 0, lwd = 0.8, add = TRUE)
lines(track1, col = "brown3")
lines(track2, col = "darkorange1")
legend("topright", legend = c("Track 1", "Track 2"), lwd = 1, col = c("brown3", "darkorange1"), pch = 1, pt.cex = 0.5, bg="white")

## Depths profiles along gps tracks
plotProfile(path1, main = "Track 1")
plotProfile(path2, main = "Track 2")

## Depth as a function of time since the deployment of gps tags
par(mar=c(5,4,1,2))
plot(track1$time, track1$depth, xlab = "Time (arbitrary unit)", ylab = "Depth (m)", type = "o", cex = 0.5, col = "brown3", xlim = range(c(track1$time, track2$time)), ylim = range(c(track1$depth, track2$depth)))
lines(track2$time, track2$depth, type = "o", cex = 0.5, col = "darkorange1")
legend("bottomright", legend = c("Track 1", "Track 2"), lwd = 1, col = c("brown3", "darkorange1"), pch = 1, pt.cex = 0.5)
```





# From the Vignette

```{r}

```
















